---
title: "Lab 2"
author: "Aleah Querns"
date: "2/3/2022"
output: html_document
---


**Load libraries:**
```{r setup, include=TRUE, message=FALSE}
library(gapminder)
library(tidyverse)
library(datasets)
library(viridis)
library(gridExtra)


```

## Question 1:
For the gapminder data order countries in reverse alphabetical order, then display the first 15. 

```{r gapminder 1, echo=TRUE}

data(gapminder)

gapminder %>%
  arrange(., desc(country)) %>%
  head(., 15)

```

**Potential issues with this process that might lead to ordering problems in a different dataset: If the chosen category wasn't listed as a factor with levels (and perhaps were a a logical statement or complex number instead) then there would probably be issues with ordering, since R wouldn't always know what order to put stuff in**


## Question 2:
Calculate and display the mean, median, and standard deviation for the Life expectancy, Population, and GDPperCapita in the gapminder dataset.

```{r gapminder 2, echo=TRUE}

#Life expectancy


LifeExpectancy<- gapminder%>%
  summarize(mean_life= mean(gapminder$lifeExp),
          median_life= median(gapminder$lifeExp),
          sd_life = sd(gapminder$lifeExp))

#Population

Population<- gapminder%>%
  summarize(mean_pop = mean(gapminder$pop),
          median_pop= median(gapminder$pop),
          sd_pop = sd(gapminder$pop))

#GDPperCapita

GDPperCap<- gapminder%>%
  summarize(mean_gdp = mean(gapminder$gdpPercap),
          median_gdp = median(gapminder$gdpPercap),
          sd_gdp = sd(gapminder$gdpPercap))

LifeExpectancy
Population
GDPperCap

```

## Question 3: 
Calculate and display in a single dataframe the mean values for Life expectancy, Population, and GDPperCapita for each continent.

```{r gapminder 3, echo=TRUE}

mean_bycontinent<- gapminder %>%
  group_by(continent)%>%
  summarize(mean_life = mean(lifeExp),
            mean_pop = mean(pop),
            mean_gdp = mean(gdpPercap))

mean_bycontinent

```

## Question 4:
Create a scatter plot of the global average Life expectancy over time and display the first 5 records for the dataset used to build this plot. Customize the display of the plot to make it less _blah_.

```{r gapminder 4, echo=TRUE}

avelife <- gapminder %>%
  group_by(year) %>%
  summarize(GlobalAvgLifeExpectancy = mean (lifeExp))

head(avelife, 5)

ggplot(avelife, aes(x=year, y=GlobalAvgLifeExpectancy)) +
  geom_point(aes(col="cyan4", size=GlobalAvgLifeExpectancy)) +
  xlab("Year") +
  ylab("Global Average Life Expectancy") +
  theme_bw()+
  theme(legend.position = "none")

```

## Question 5:
Pull in a dataset of your choice from the datasets package that has a dataframe structure. Then examine the structure of the data including: class types, header information, and dimensions.

**Here, I used the beavers dataset (specifically beaver1)**


```{r random data 1, echo=TRUE}
data(beavers) #focusing on beaver 1

str(beaver1)

```

## Question 6:
Calculate general summary statistics for the dataset.

```{r random data 2, echo=TRUE}

beaver1_summary<-
  beaver1 %>%
  summarize(., ave_temp = mean(temp),
            sd_temp= sd(temp),
            times_active = sum(activ),
            num_obs = length(time))

beaver1_summary

```

## Question 7:
Calculate summary statistics for groups/subsets of the dataset (these groupings could be based on time period, species, location, etc…).

```{r random data 3, echo=TRUE}

beaver1_summary_byday<-
  beaver1 %>%
  group_by(day)%>%
  summarize(., ave_temp = mean(temp),
            sd_temp = sd(temp),
            times_active = sum(activ),
            num_obs = length(time))

beaver1_summary_byday

```

## Question 8:
Create 2 plots of your choice, one showing grouped data and the other showing ungrouped data. Give each plot a title and descriptive x and y axis labels. Customize at least one graphic option (color, point type, line width, etc…) on these plots.

**Wasn't really sure how to graph the previously described summary statistics in a way that would be meaningful, so instead I made a histogram (ungrouped) and time series/line graph (grouped by day)**

```{r random data 4, echo=TRUE}

beaver1$day <- as.character(beaver1$day)

 p1<-ggplot(data=beaver1, aes(temp)) +
    geom_histogram(fill="cyan4", col="black", bins=50)+
   xlab("Temperature (degrees Celsius)") +
   ylab("Count")+
   theme_bw()
  #ungrouped histogram
 
 p2<-ggplot(data=beaver1, aes(x=time, y=temp))+
   geom_line(aes(group=day, col=day))+
   geom_point(aes(group=day, col=day))+
   xlab("Time (military)")+
   ylab("Temperature (degrees Celsius)") +
   theme_bw() +
   scale_color_viridis(discrete=TRUE, option = "D")
 
 grid.arrange(p1,p2)
 

```