---
title: "Lab 5"
author: "Aleah Querns"
date: "2/21/2022"
output: html_document
---

# The Setup

```{r setup, include=TRUE, message=FALSE}
library(tidyverse)
library(sf)
library(spData)
library(gridExtra)

knitr::opts_knit$set(root.dir = '~/Documents/K-State Year 2/GEOG728/GEOG728_GIT')
```

# The Dataset

```{r dataset, echo=TRUE}

data(world)
data(worldbank_df)

world
worldbank_df

```
###Q1: Examine the world and worldbank_df datasets. Describe what kind of data objects they are and their general structure and makeup. Then combine these two data objects into a single spatial data object. Does the resulting data object look right?

```{r yeet, echo=TRUE}
class(world)
class(worldbank_df)

str(world)
str(worldbank_df)

setdiff(world$name_long, worldbank_df$name)

worldbank_df$name<-gsub(pattern = "Cote d'Ivoire", replacement= "Côte d'Ivoire", worldbank_df$name)

worldbank_df$name<-gsub(pattern = "Swaziland", replacement= "eSwatini", worldbank_df$name)

worldbank_df$name<-gsub(pattern = "Republic of Congo", replacement= "Republic of the Congo", worldbank_df$name)


worldfull<-merge(world, worldbank_df, by.x ="name_long", by.y= "name", na.rm=F)

colSums(is.na(worldfull)) 

```

_"world"_

+ Data type: "world" is an "sf", "tbl_df", "tbl", and "data.frame"

+ General structure: 177 rows, 11 columns. First 6 columns (characters) provide various types of country-related descriptions for data entries. Columns 7-9 (numeric) provide various stats about these countries (pop=population; lifeExp= life expectancy; gdpPercap= GDP per capita), and the final column "geom" ("sfc_MULTIPOLYGON") provides polygons for each country.


_"worldbank_df"_

+ Data type: "worldbank_df" is a "tbl_df", "tbl", and "data.frame"

+ General structure: 177 rows, 7 columns. First two columns (characters) provide names for countries. Columns 5-7 (numeric) provide various stats on each country (HDI, urban populatin size, unemployment, population growth, literacy rates)

**At first, the datasets weren't merging correctly due to small differences in country names between the two. To correct this, I used gsub to correct the errors in the worldbank_df**

### Q2: Make some basic (non-spatial) plots to explore differences in urban population and literacy across countries.

```{r basic plots 1, echo=TRUE}

worldfull%>%
  filter(!is.na(literacy), literacy !=0)%>%
ggplot(aes(x=reorder(name_long, literacy), y=literacy)) +
  geom_bar(stat = 'identity')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  xlab ("Country") +
  ylab("Adult Literacy Rate")

worldfull%>%
  filter(!is.na(urban_pop), urban_pop !=0)%>%
ggplot(aes(x=reorder(name_long, urban_pop), y=urban_pop)) +
  geom_bar(stat = 'identity')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  xlab ("Country") +
  ylab("Urban Population Size")

```

_As always, these plots are hard to view because of the number of countries, so it's best to view them in another window_

### Q3: Now make some maps to explore urban population and literacy across countries.

``` {r maps 1}

ggplot(worldfull)+
  geom_sf(aes(fill=log(urban_pop))) +
  theme_bw()+
  guides(fill=guide_legend(title="log(Urban Population Size)"))

ggplot(worldfull)+
  geom_sf(aes(fill=literacy)) +
  theme_bw()+
  guides(fill=guide_legend(title="Adult Literacy Rate"))

```

_Interestingly, it seems like most countries aren't providing data on literacy rates_

### Q4: Make some basic (non-spatial) plots to explore differences in urban population and literacy across continents.

```{r basic plots 2, echo=TRUE}

worldfull%>%
  filter(!is.na(literacy), literacy !=0)%>%
  group_by(continent)%>%
  summarise(literacy_mean = mean(literacy)) %>%
ggplot(aes(x=reorder(continent, literacy_mean), y=literacy_mean)) +
  geom_bar(stat = 'identity')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  xlab ("Continent") +
  ylab("Adult Literacy Rate")

worldfull%>%
  filter(!is.na(urban_pop), urban_pop !=0)%>%
  group_by(continent)%>%
  summarise(urban_pop_mean = mean(urban_pop)) %>%
ggplot(aes(x=reorder(continent, urban_pop_mean), y=urban_pop_mean)) +
  geom_bar(stat = 'identity')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  xlab ("Continent") +
  ylab("Urban Population Size")

```

### Q5: Now make some maps to explore differences in urban population and literacy across continents. Change to a different, projected coordinate system for these maps.

``` {r maps 2}
worldfullproj<-st_make_valid(st_transform(worldfull, "+proj=natearth +lon_0=1 +datum=WGS84 +units=m +no_defs"))

worldfullproj %>%
  group_by(continent) %>%
  summarise(urban_pop_mean = mean(urban_pop, na.rm=TRUE)) %>%
ggplot()+
  geom_sf(aes(fill=log(urban_pop_mean))) +
  theme_bw()+
  guides(fill=guide_legend(title="log(Mean Urban Population Size Across Continent)"))+
  theme(legend.title = element_text(size = 8))

worldfullproj %>%
  group_by(continent) %>%
  summarise(lit_mean = mean(literacy, na.rm=TRUE)) %>%
ggplot()+
  geom_sf(aes(fill=lit_mean)) +
  theme_bw()+
  guides(fill=guide_legend(title="Mean Adult Literacy Rate Across Continent"))+
  theme(legend.title = element_text(size = 8))

```
